An elaborate data set on human gait and the effect of mechanical perturbations

### 2014-12-13 paper.tex
Proofed the abstract and introduction.

  Here we share a rich gait data set collected from fifteen subjects walking at
  three speeds on an instrumented treadmill. Each trial consists of 120 seconds
  of normal walking and 480 seconds of walking while being longitudinally
  perturbed during each stance phase with pseudo-random
  fluctuations in the speed of the treadmill belt. A total of approximately 1.5
  hours of normal walking ($>5000$ gait cycles) and 6 hours of perturbed walking
  ($>20,000$ gait cycles) is included in the data set. We provide full body
  marker trajectories and ground reaction loads in addition to a presentation of
  processed data that includes gait events, 2D joint angles, angular rates,
  and joint torques and the software used for the computations. The protocol is
  described in detail along with additional elaborate meta data. This data can
  likely be useful for validating or generating mathematical models that are
  capable of simulating normal periodic gait and non-periodic, perturbed gaits.

@@

\section*{Introduction}
%
The collection of dynamical data during human walking has a long history
beginning with the first motion pictures and now with modern marker based
motion capture techniques, image and video analysis, and high fidelity ground
reaction load measurements. Even though years of data on thousands of subjects
now exist, this data is not widely disseminated, well organized, nor available
with few or no restrictions. David Winter's published normative gait
data~\cite{Winter1990} is widely used in biomechanics studies, yet it comes
from relatively few subjects and only a small number of gait cycles per
subject. This small source has successfully inspired powered prosthetic control
design~\cite{Sup2008}, but success in other fields using large sets of
data for discovery lead one to believe that more elaborate data sets may 
benefit the field of human motion studies.
To enable such work, data needs
to be shared extensively, organized, and curated to enabled analysts.

There are some notable gait data sets and databases besides Winter's authoritative
set that are publicly available. Notably the International Society of
Biomechanics has maintained a web page (http://isbweb.org/data) since
approximately 1995 that includes data sets for download and most unencumbered
use. Vaughn, et. al's data~\cite{Vaughan1992} with kinematics and force plate
measurement from several subjects. Chris Kirtley shares normative gait from
several studies on the CGA Normative Gait Database~\cite{Kirtley2014} and these
files have influenced and been included in other studies, for
example the average gait cycles from children used in~\cite{Bogert2003}.

Chester et. al~\cite{Chester2007} report on a large gait database comparison
where one database contained kinematic data of 409 gait cycles of children from
1 to 7 years old but the data does not seem to be publicly available. Tirosh
et. al, recognized the need of a comprehensive data base for clinical gait data
and created the Gaitabase~\cite{Tirosh2010}.  This database may contain a
substantial amount of data but it is encumbered by a very complicated and
restrictive license and sharing scheme. The University of Wisconsin at LaCrosse
has a normative gait data set~\cite{Wilson2014} from 25 subjects with lower
extremity marker data and force plate measurements from a single stance phase (TvdB: gait cycle?).

More recent examples of biomechanists sharing their data alongside
publications are: \cite{Bogert2013} which includes processed joint kinematics
from eleven subjects walking for few gait cycles and \cite{Wang2014} who
includes data from 10 subjects walking for five minutes each at five different
speeds but only a small set of lower extremity markers are present. The second
is notable because it publishes the data in Dryad, a modern citable data
repository.

The publicly available gait data is small compared to the number of gait
studies that have been performed over the years. And the data that is available
in general suffers from things such as few subjects, few gait cycles, few
markers, no raw data, limited force plate measurements, lack of meta data, and
restrictive licensing. To help with this situation we are making the data we
collected for our research purposes publicly available and free of the
previously mentioned deficiencies. Not only do we provide a larger set of
normative gait data that has been previously available we include an even large
set of data in which the subject is being perturbed which does not exist at
all. We believe both of these sets of data can serve a variety of use cases and
hope that we can save time for researchers with use cases for this data.

Our use case for the data is centered around emerging powered prosthetics and
orthotics need of bio-inspired control systems. Ideally, a powered prosthetic
would behave in such a way that the user would feel like their limb was never
disabled. There are a variety of approaches to developing bio-inspired control
systems some of which aim to mimic the reactions and motion of an able-bodied
person. A modern gait lab is able to collect a variety of kinematic, kinetic,
and physiological data from humans during gait. This data can potentially be
used to drive the design of the human-mimicking controller. With a rich enough
data set, one may be able to identify control mechanisms used during a human's
natural gait and recovery from perturbations. We have collected data that we
believe is richer than previous gait data sets and may be rich enough for
control identification.  The data can also be used for verification purposes
for controllers that have been designed in other manners.

With all of this in mind, we collected over five and half hours of gait data
from fifteen able bodied subjects and includes over 25,000 gait cycles. Most of
the subjects walked at three different speeds on an instrumented treadmill
while we collected full body marker locations and ground reaction loads from a
pair of force plates. The protocol for the majority of the trials includes two
minutes of normal walking and eight minutes of walking under the influence of
pseudo-random belt speed fluctuations. The data has been organized complete with
rich meta data and made available in the most unrestrictive form for other
research uses following modern best practices in data sharing~\cite{White2013}.

>>>>

  Here we share a rich gait data set collected from fifteen subjects walking at
  three speeds on an instrumented treadmill. Each trial consists of 120 seconds
  of normal walking and 480 seconds of walking while being longitudinally
  perturbed during each stance phase with pseudo-random fluctuations in the
  speed of the treadmill belt. A total of approximately 1.5 hours of normal
  walking ($>5000$ gait cycles) and 6 hours of perturbed walking ($>20,000$
  gait cycles) is included in the data set. We provide full body marker
  trajectories and ground reaction loads in addition to a presentation of
  processed data that includes gait events, 2D joint angles, angular rates, and
  joint torques along with the open source software used for the computations.
  The protocol is described in detail and supported with additional elaborate
  meta data for each trial. This data can likely be useful for validating or
  generating mathematical models that are capable of simulating normal periodic
  gait and non-periodic, perturbed gaits.

@@

\section*{Introduction}
%
The collection of dynamical data during human walking has a long history
beginning with the first motion pictures and now with modern marker based
motion capture techniques and high fidelity ground reaction load measurements.
Even though years of data on thousands of subjects now exist, this data is not
widely disseminated, well organized, nor available with few or no restrictions.
David Winter's published normative gait data~\cite{Winter1990} is widely used
in biomechanical studies, yet it comes from relatively few subjects and only a
small number of gait cycles per subject. This small source has successfully
inspired many other studies, such as powered prosthetic control
design~\cite{Sup2008}, but success in other research fields using large sets of
data for discovery lead one to believe that more elaborate data sets may
benefit the field of human motion studies. To enable such work, biomechanical
data needs to be shared extensively, organized, and curated to enabled
future analysts.

There are some notable gait data sets and databases besides Winter's
authoritative set that are publicly available. The International Society of
Biomechanics has maintained a web page (http://isbweb.org/data) since
approximately 1995 that includes data sets for download and mostly unencumbered
use. For example, Vaughn, et. al's data~\cite{Vaughan1992} with kinematics and
force plate measurement from several subjects is available on the site. At
another website, the CGA Normative Gait Database~\cite{Kirtley2014}, Chris
Kirtley shares normative gait data from several studies and these files have
influenced other studies, for example the average gait cycles from children
used in~\cite{Bogert2003}.

Chester et. al~\cite{Chester2007} report on a large gait database comparison
where one database contained kinematic data of 409 gait cycles of children from
1 to 7 years old but the data does not seem to be publicly available. This is
unfortunately typical. But Tirosh et. al, recognized the need for a
comprehensive data base for clinical gait data and created the
Gaitabase~\cite{Tirosh2010}. This database may contain a substantial amount of
data but it is encumbered by a very complicated and restrictive license and
sharing scheme. However, there are examples of data with less restrictions. The
University of Wisconsin at LaCrosse has an easily accessible normative gait
data set~\cite{Wilson2014} from 25 subjects with lower extremity marker data
from multiple gait cycles and force plate measurements from a single gait
cycle.

More recent examples of biomechanists sharing their data alongside publications
are: \cite{Bogert2013} which includes full body joint kinematics and kinetics
from eleven subjects walking for a small number of gait cycles and
\cite{Wang2014} who includes a larger set of data from ten subjects walking for
five minutes each at three different speeds but only a small set of lower
extremity markers are present. The second is notable because it publishes the
data in Dryad, a modern citable data repository.

The publicly available gait data is small compared to the number of gait
studies that have been performed over the years. The data that is available
generally suffers from limitations such as few subjects, few gait cycles, few
markers, highly clinical, no raw data, limited force plate measurements, lack
of meta data, non-standard formats, and restrictive licensing. To help with
this situation we are making the data we collected for our research purposes
publicly available and free of the previously mentioned deficiencies. Not only
do we provide a larger set of normative gait data that has been previously
available, we also include an even larger set of data in which the subject is
being perturbed, something that does not currently exist. We believe both of
these sets of data can serve a variety of use cases and hope that we can save
time and effort for future researchers by sharing it.

Our use case for the data is centered around the need of bio-inspired control
systems for emerging powered prosthetics and orthotics. Ideally, a powered
prosthetic would behave in such a way that the user would feel like their limb
was never disabled. There are a variety of approaches to developing
bio-inspired control systems, some of which aim to mimic the reactions and
motion of an able-bodied person. A modern gait lab is able to collect a variety
of kinematic, kinetic, and physiological data from humans during gait. This
data can potentially be used to drive the design of the human-mimicking
controller. With a rich enough data set, one may be able to identify control
mechanisms used during a human's natural gait and recovery from perturbations.
We have collected data that is richer than previous gait data sets and may be
rich enough for control identification. The data can also be used for
verification purposes for controllers that have been designed in other manners.

With all of this in mind, we collected over seven and half hours of gait data
from fifteen able bodied subjects which amounts to over 25,000 gait cycles. The
subjects walked at three different speeds on an instrumented treadmill while we
collected full body marker locations and ground reaction loads from a pair of
force plates. The protocol for the majority of the trials included two minutes
of normal walking and eight minutes of walking under the influence of
pseudo-random belt speed fluctuations. The data has been organized complete
with rich meta data and made available in the most unrestrictive form for other
research uses following modern best practices in data sharing~\cite{White2013}.
